Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-07-17 manuscript/revision2/revision2.qmd
revision2: Sync cross-references between revision, SI Appendix, and response letter.
Also, add Figures to response letter and change templating to allow this

Plant phenological timings are major components of plant fitness affected by multiple external environmental cues (e.g. degree of winter chilling, day length, temperature, and water availability) that signal existing or upcoming growing conditions [@bauerle_photoperiodic_2012; @andres2012genetic; @korner2010phenology; @maple2024]. Genetic responses to environmental cues determine the speed, timing, and energy apportioned to vegetative and reproductive growth and shape both the individual's lifespan and its lifetime production of viable seed. Day length (or photoperiod) is one of the most predictable environmental cues, and genetic sensitivity to photoperiod protects plants from potentially fatal consequences of phenological responses to temperature cues at the "wrong" time of year. However, the usefulness of specific environmental cues depends on both features of the environment, such as cue predictability and relevance, and the species' adaptive strategies @botero2015evolutionary. Species with wide natural distributions can have multiple distinct environmentally cued phenological responses: for example, populations of sunflower (*Helianthus annuus*) exhibit day-neutral, facultative short day, and facultative long-day flowering responses, which vary with their environments [@blackman2013interacting; @henry2014transitions]. Distinct genetic responses in different environments are known as genotype-by-environment interactions, or GxE.

Flowering time, or the transition from vegetative to reproductive growth, is a common subject of GxE research [@blackman2013interacting; @henry2014transitions; @aagren2017adaptive; @brachi2010linkage; @dittmar2014flowering; @li2018genomic; @romero2017study; @bustos-korts2019], a key output of selection driving adaptation to local environments [@andres2012genetic; @wadgymar2017identifying; @blumel2015flowering; @fournier-level2016], and a selection target for crop improvement to adapt crops to local or future environments @jung2009flowering. Changing flowering responsiveness to photoperiod cues has allowed geographic range expansion and increased yields in several cereal species [@turner2005pseudo; @faure2012mutation; @hung2012zmcct; @zakhrabekova2012induced; @yang2013oself3] and other crops [@pin2012multifaceted; @weller2019parallel]. Recent statistical advances in studying phenological GxE have involved determining critical environmental indices before the phenological event occurs, such as photothermal time within a critical growth window @li2018genomic. However, most studies of flowering GxE focus on finding a single, best fitting form of genotype-environment covariance, despite the key expectation that different genetic subpopulations, and even different genomic regions, have likely evolved distinct patterns of GxE [@Weine2023.06.21.545998]. Additionally, despite theoretical predictions that local adaptation should involve antagonistic pleiotropy, or alleles with effects with opposing fitness outcomes [@levene1953genetic; @felsenstein1976theoretical; @kawecki2004conceptual; @hedrick1986genetic], previous empirical work has found limited evidence of antagonistic pleiotropy [@des2013genotype; @wadgymar2017identifying; @millet2016]. However, this work has been limited by a known statistical bias that reduced detection of genetic effects that differ in sign [@des2013genotype; @anderson2013genetic; @anderson2011evolutionary]. Thus, despite substantial interest in the frequencies of various forms of GxE, the frequency of sign-changing GxE relative to other forms of GxE remains unknown.

Previous research suggests that switchgrass phenological timings should have GxWeather and that these timings could differ by genetic subpopulation. Switchgrass is considered a short-day plant with reproductive development strongly linked to day of the year @mitchell1997predicting. However, as part of its wide environmental adaptation across the eastern half of North America, its photoperiodicity has been predicted to differ by plant latitude of origin [@parrish2005biology; @casler2004latitudinal]. We previously found divergent Midwest and Gulf genetic subpopulations of switchgrass with distinct sets of environmental adaptations, in that both populations had distinct genetic variation associated with each of two fitness proxies, biomass and overwinter survival @lovell2021genomic. The Midwest genetic subpopulation is primarily composed of individuals from the well-studied upland switchgrass ecotype [@porter1966analysis; @milano2016genetic], while the Gulf subpopulation has individuals from the well-studied lowland ecotype and the phenotypically intermediate coastal ecotype @lovell2021genomic.

@@

We expected to observe common rank-changing GxE at the level of individual loci as this is a key theoretical prediction of local adaptation [@felsenstein1976theoretical; @levene1953genetic; @kawecki2004conceptual; @hedrick1986genetic]. To determine the frequency of rank-changing GxE, we used the local false sign rate (lfsr), an analogue of the lfdr that establishes confidence in the effect sign, not the effect's difference from zero, to determine significance. We required lfsr significance (p \< 0.05) in both gardens to include effects. This means that our tests for a sign change between gardens carry an equal statistical burden to those for effects with the same sign. We separated kinds of effects at the level of individual loci into SNP effects that differ in sign between gardens (effects with rank-changing GxE between gardens), SNP effects that differ in magnitude (effects that are large in one garden or region and smaller in others), and SNP effects that are indistinguishable in two gardens (similarly large or small in comparisons between gardens) (@fig-effects A).

@@

The formation and resequencing of the diversity panel has been described previously @lovell2021genomic. The diversity panel contained 134 sequenced, clonally propagated individuals from the Midwest genetic subpopulation, and 229 from the Gulf genetic subpopulation. To allow for the possibility that different subpopulations had different strengths of connection between our phenotypes and genotypes [@kortefarlow2013], we conducted three sets of genetic analyses: on Gulf and Midwest genotypes separately, and on both subpopulations together ('Both' subpopulations). Analyses to determine narrow-sense heritability (h^2^) for green-up and flowering were done using linear mixed models and followed @lovell2021genomic (SI Appendix, Section S6).

@@

Using the lfsr rather than the local false discovery rate (lfdr) is a critical change in our ability to detect alleles directly contributing to rank-changing GxE between environments. The lfdr, like other measures of FDR, focuses on if we have enough evidence to reject the null hypothesis that an effect j is 0, or that there is a significant effect. Previous studies of antagonistic pleiotropy (e.g. [@lowry2019qtl]) have used the lfdr or equivalent statistical tests to detect antagonistic pleiotropy. These tests were conservative, in that they required two non-zero effects of different signs, while tests for differential sensitivity required only one non-zero effect. This previous work recognized that this testing bias could lead to undercounting occurrences of antagonistic pleiotropy [@anderson2011evolutionary; @des2013genotype], and sought to reduce it by permutation [@anderson2013genetic]. However, using the lfsr to test for allelic effects that differ in sign does not undercount these occurences, as this statistic answers a fundamentally different question. For each effect $j$, the $lfsr_j$ is defined as the probability that we make an error in the sign of effect $j$ if we were forced to declare the effect positive or negative @10.1093/biostatistics/kxw041. Thus, rather than asking "Are these two effects different?" - as we reasonably expect two effects to be, even if this difference cannot be measured - the local false sign rate answers a more meaningful question: Can we be confident in the sign of this effect?

>>>>

Plant phenological timings are major components of plant fitness affected by multiple external environmental cues (e.g. degree of winter chilling, day length, temperature, and water availability) that signal existing or upcoming growing conditions [@bauerle_photoperiodic_2012; @andres2012genetic; @korner2010phenology; @maple2024]. Genetic responses to environmental cues determine the speed, timing, and energy apportioned to vegetative and reproductive growth and shape the individual's physiological responses, lifespan, and lifetime production of viable seed. Day length (or photoperiod) is one of the most predictable environmental cues, and genetic sensitivity to photoperiod protects plants from potentially fatal consequences of phenological responses to temperature cues at the "wrong" time of year. However, the usefulness of specific environmental cues depends on both features of the environment, such as cue predictability and relevance, and the species' adaptive strategies @botero2015evolutionary. Species with wide natural distributions can have multiple distinct environmentally cued phenological responses: for example, populations of sunflower (*Helianthus annuus*) exhibit day-neutral, facultative short day, and facultative long-day flowering responses, which vary with their environments [@blackman2013interacting; @henry2014transitions]. Distinct genetic responses in different environments are known as genotype-by-environment interactions, or GxE.

Flowering time, or the transition from vegetative to reproductive growth, is a common subject of GxE research [@blackman2013interacting; @henry2014transitions; @aagren2017adaptive; @brachi2010linkage; @dittmar2014flowering; @li2018genomic; @romero2017study; @bustos-korts2019], a key output of selection driving adaptation to local environments [@unterseer2016; @andres2012genetic; @wadgymar2017identifying; @blumel2015flowering; @fournier-level2016], and a selection target for crop improvement to adapt crops to local or future environments @jung2009flowering. Changing flowering responsiveness to photoperiod cues has allowed geographic range expansion and increased yields in several cereal species [@unterseer2016; @turner2005pseudo; @faure2012mutation; @hung2012zmcct; @zakhrabekova2012induced; @yang2013oself3] and other crops [@pin2012multifaceted; @weller2019parallel]. Recent statistical advances in studying phenological GxE have involved determining critical environmental indices before the phenological event occurs, such as photothermal time within a critical growth window @li2018genomic. However, most studies of flowering GxE focus on finding a single, best fitting form of genotype-environment covariance, despite the key expectation that different genetic subpopulations, and even different genomic regions, have likely evolved distinct patterns of GxE [@Weine2023.06.21.545998]. Additionally, despite theoretical predictions that local adaptation should involve antagonistic pleiotropy, or alleles with effects with opposing fitness outcomes [@levene1953genetic; @felsenstein1976theoretical; @kawecki2004conceptual; @hedrick1986genetic], previous empirical work has found limited evidence of antagonistic pleiotropy [@des2013genotype; @wadgymar2017identifying; @millet2016]. However, this work has been limited by a known statistical bias that reduced detection of genetic effects that differ in sign [@des2013genotype; @anderson2013genetic; @anderson2011evolutionary]. Thus, despite substantial interest in the frequencies of various forms of GxE, the frequency of sign-changing GxE relative to other forms of GxE remains unknown.

Previous research suggests that switchgrass phenological timings should have GxWeather and that these timings could differ by genetic subpopulation. Switchgrass is considered a short-day plant with reproductive development strongly linked to day of the year @mitchell1997predicting. However, as part of its wide environmental adaptation across the eastern half of North America, its photoperiodicity has been predicted to differ by plant latitude of origin [@parrish2005biology; @casler2004latitudinal]. We previously found divergent Midwest and Gulf genetic subpopulations of switchgrass with distinct sets of environmental adaptations, in that both populations had distinct genetic variation associated with each of two fitness proxies, biomass and overwinter survival @lovell2021. The Midwest genetic subpopulation is primarily composed of individuals from the well-studied upland switchgrass ecotype [@porter1966analysis; @milano2016genetic], while the Gulf subpopulation has individuals from the well-studied lowland ecotype and the phenotypically intermediate coastal ecotype @lovell2021.

@@

We expected to observe common rank-changing GxE at the level of individual loci as this is a key theoretical prediction of local adaptation [@felsenstein1976theoretical; @levene1953genetic; @kawecki2004conceptual; @hedrick1986genetic]. To determine the frequency of rank-changing GxE, we used the local false sign rate (lfsr), an analogue of the local false discovery rate that establishes confidence in the effect sign, not the effect's difference from zero, to determine significance. We required lfsr significance (p \< 0.05) in both gardens to include effects. This means that our tests for a sign change between gardens carry an equal statistical burden to those for effects with the same sign. We separated kinds of effects at the level of individual loci into SNP effects that differ in sign between gardens (effects with rank-changing GxE between gardens), SNP effects that differ in magnitude (effects that are large in one garden or region and smaller in others), and SNP effects that are indistinguishable in two gardens (similarly large or small in comparisons between gardens) (@fig-effects A).

@@

The formation and resequencing of the diversity panel has been described previously @lovell2021. The diversity panel contained 134 sequenced, clonally propagated individuals from the Midwest genetic subpopulation, and 229 from the Gulf genetic subpopulation. To allow for the possibility that different subpopulations had different strengths of connection between our phenotypes and genotypes [@kortefarlow2013], we conducted three sets of genetic analyses: on Gulf and Midwest genotypes separately, and on both subpopulations together ('Both' subpopulations). Analyses to determine narrow-sense heritability (h^2^) for green-up and flowering were done using linear mixed models and followed @lovell2021 (SI Appendix, Section S6).

@@

Using the lfsr rather than the local false discovery rate (local FDR) is a critical change in our ability to detect alleles directly contributing to rank-changing GxE between environments. The local FDR, like other measures of FDR, focuses on if we have enough evidence to reject the null hypothesis that an effect j is 0, or that there is a significant effect. Previous studies of antagonistic pleiotropy (e.g. [@lowry2019qtl]) have used the local FDR or equivalent statistical tests to detect antagonistic pleiotropy. These tests were conservative, in that they required two non-zero effects of different signs, while tests for differential sensitivity required only one non-zero effect. This previous work recognized that this testing bias could lead to undercounting occurrences of antagonistic pleiotropy [@anderson2011evolutionary; @des2013genotype], and sought to reduce it by permutation [@anderson2013genetic]. However, using the lfsr to test for allelic effects that differ in sign does not undercount these occurences, as this statistic answers a fundamentally different question. For each effect $j$, the $lfsr_j$ is defined as the probability that we make an error in the sign of effect $j$ if we were forced to declare the effect positive or negative @10.1093/biostatistics/kxw041. Thus, rather than asking "Are these two effects different?" - as we reasonably expect two effects to be, even if this difference cannot be measured - the local false sign rate answers a more meaningful question: Can we be confident in the sign of this effect?
